---
title: "EDA Assignmemnt2"
author: "Pamela Afful"
format: pdf
editor: visual

execute: 
  cache: true
  warning: false
  message: false
  echo: false
---

```{r, read_lib, include=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(zoo)
library(plotly)
```

# Data Wrangling

Read in the files

```{r,read_data,echo=FALSE,eval=TRUE}

query=read.csv("query.csv")
earth_quakes=read.csv("Earthquakes_1965_2016.csv")


```

Check Dataset Dimensions

```{r,data_det,echo=FALSE,eval=TRUE}
# check query df details
dim(query)
dim(earth_quakes)
```

Convert time characters to date variables for both dataframes

Create truncated date variables, only including the date part

```{r,date_con,echo=FALSE,eval=TRUE}
# Change date columns to datetime
query$dt_date_time=ymd_hms(query$time)
query$date_final=date(query$dt_date_time)

```

Earthquakes data

```{r date_con_2, eval=TRUE, echo=FALSE}
#| message: true
#| warning: false
#| fig-height: 6
#| fig-width: 6
#| paged-print: true
# Split data into 2 then put back together

# earth_quakes_final=earth_quakes%>% mutate(dt_date_time= mdy_hms(paste(Date,Time)))
# earth_quakes_final$date_final=date(earth_quakes_final$dt_date_time)


earth_quakes$dt_date_time=earth_quakes%>% mutate(dt_date_time= mdy_hms(paste(Date,Time))) %>% pull(dt_date_time)

earth_quake_na=earth_quakes %>% filter(is.na(dt_date_time))
earth_quake_non_na=earth_quakes %>% filter(!is.na(dt_date_time))
earth_quake_na$dt_date_time=earth_quake_na %>% mutate(dt_date_time=ymd_hms(Date)) %>% pull(dt_date_time) 

earth_quakes_final=rbind(earth_quake_na,earth_quake_non_na)
earth_quakes_final$date_final=date(earth_quakes_final$dt_date_time)

```

Transform column names to lower case for both dataframes

Also, remove the dot character, "." from the column names of the earthquake data and lower the case of the values in the id column for the earthquakes data

```{r,lower_col,echo=FALSE,eval=TRUE}
# lower the case of column names
colnames(query)=tolower(colnames(query))


# lower the case of the columns names in earth quake df and remove period character in column names
colnames(earth_quakes_final)=tolower(colnames(earth_quakes_final))
colnames(earth_quakes_final)=gsub("\\.","",colnames(earth_quakes_final))

# lower the case of the ID values
earth_quakes_final$id=tolower(earth_quakes_final$id)
earth_quakes_final$type=tolower(earth_quakes_final$type)

names(earth_quakes_final)
names(query)
```

Finally join the dataframes on common column values

```{r,df_join,echo=FALSE,eval=TRUE}

# find common columns in both dataframes
# these will be used to join the two dataframes
eq_cols=names(earth_quakes_final)
query_cols=names(query)

join_cols=intersect(eq_cols,query_cols)
join_cols=join_cols[join_cols!='time']

ex_cols=c('date_final','id','type')

# rename common cols in the query dataframe to get rid of duplicated column names once the two dataframes are joined
for(col in join_cols){
  if(col %in% ex_cols){
  names(query)[names(query) == col] = col}
  else{
    names(query)[names(query) == col] = paste0(col,'_query')
  }
    
    
}

df_=full_join(earth_quakes_final,query, by=ex_cols )

# filter for earthqauke data

df_earthquake=dplyr::filter(df_, `type` =='earthquake')

# Fill in na for some columns


df_earthquake=df_earthquake %>% mutate(magnitude=case_when(is.na(magnitude) & !is.na(mag)~ mag,
                                             TRUE ~magnitude
                                             )) %>% mutate(rootmeansquare=case_when(is.na(rootmeansquare) & !is.na(rms)~ rms,
                                             TRUE ~rootmeansquare
                                             )) %>% 
mutate(rootmeansquare=case_when(is.na(magnitudeerror) & !is.na(magerror)~ magerror,
                                             TRUE ~magnitudeerror
                                             )) %>% 
  
mutate(depth=case_when(is.na(depth) & !is.na(depth_query)~ depth_query,
                                             TRUE ~depth
                                             )) %>% 
 
mutate(depth=case_when(is.na(deptherror) & !is.na(deptherror_query)~ deptherror_query,
                                             TRUE ~deptherror
                                             ))  %>% 
mutate(depth=case_when(is.na(dt_date_time) & !is.na(dt_date_time_query)~ dt_date_time_query,
                                             TRUE ~dt_date_time
                                             ))  %>%  
  
mutate(depth=case_when(is.na(horizontalerror) & !is.na(horizontalerror_query)~ horizontalerror_query,
                                             TRUE ~horizontalerror
                                             )) %>% 
mutate(depth=case_when(is.na(latitude) & !is.na(latitude_query)~ latitude_query,
                                             TRUE ~latitude
                                             )) %>% 
mutate(depth=case_when(is.na(longitude) & !is.na(longitude_query)~ longitude_query,
                                             TRUE ~longitude
                                             )) %>% 
mutate(depth=case_when(is.na(longitude) & !is.na(longitude_query)~ longitude_query,
                                             TRUE ~longitude
                                             )) %>% 
  
mutate(depth=case_when(is.na(source) & !is.na(magsource)~ magsource,
                                             TRUE ~source
                                             )) %>% 
  
mutate(depth=case_when(is.na(status) & !is.na(status_query)~ status_query,
                                             TRUE ~status
                                             )) 
   


#sort(names(df_earthquake))

```

### Creating Magnitude Scale

```{r,mag_cat,echo=FALSE,eval=TRUE}
df_earthquake$richter_scale=df_earthquake %>% mutate(richter_scale=case_when(
                                                  #magnitudetype=='ML' &
                                                  magnitude>=1 & magnitude<=1.9~'Micro',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=2 & magnitude<=2.9~'Minor',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=3 & magnitude<=3.9~'Slight',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=4 & magnitude<=4.9~'Light',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=5 & magnitude<=5.99~'Moderate',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=6 & magnitude<=6.9~'Strong',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=7 & magnitude<=7.9~'Major',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=8 & magnitude<=8.9~'Great',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=9 & magnitude<=9.9~'Extreme'
                                                  #magnitudetype!='ML'~'Other',
                                                  
                                                  
                                                  
                                                    
)
                         ) %>% pull(richter_scale)

```

# General Data Analysis

Find count/frequency of earth quake types per magnitude

```{r,mg_summ}
# Magnitude summary Statistics
summary(df_earthquake$magnitude)
```

```{r,bar_plot,echo=FALSE,eval=TRUE}

p1=df_earthquake %>%  
  distinct(id,richter_scale) %>% 
  mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  ggplot(aes(x=richter_scale))+
  geom_bar()+
  labs(title = "Counts of Earthquake Categories",
       x = "magnitude category") +
  theme_minimal()

# df_earthquake %>%  
#   select(magnitude,richter_scale) %>% 
#   ggplot(aes(x=magnitude,fill=richter_scale))+
#   geom_histogram(position = "identity", alpha = 0.6, binwidth = 0.5)+
#   labs(title = "Earth Quake Magnitudes Per Richter Scale Category (All Magnitude Types)",
#        x = "magnitude") +
#   theme_minimal()

p2=df_earthquake %>%  
  select(magnitude,richter_scale) %>% 
  ggplot(aes(y=magnitude))+
  geom_boxplot()+
  labs(title = "Distribution of Earthquake Magnitudes",
       y = "magnitude") +
  theme_minimal()

p1+p2
  
```

```{r,depth_mag,echo=FALSE,eval=TRUE}

p1=df_earthquake %>%  
  select(magnitude,depth,richter_scale) %>% 
    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  ggplot(aes(x=richter_scale,y=depth))+
  geom_boxplot()+
  labs(title = "Depth vs Earthquake Category",
       y = 'earth quake depth') +
  theme_minimal()


p2=df_earthquake %>% filter(!is.na(depth)) %>% 

    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  group_by(richter_scale) %>% 
  summarise(mean_depth=mean(depth)) %>% 
  
  ggplot(aes(x=richter_scale,y=mean_depth))+
  geom_col(stat = 'identity')+
  labs(title = "Average horizontal distance per earthquake category",
       y = "depth") +
  theme_minimal()

p1/p2

```

```{r dist_mag, eval=TRUE, echo=FALSE}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 12
#| paged-print: false

p1=df_earthquake %>% filter(!is.na(horizontaldistance)) %>% 

    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  ggplot(aes(x=horizontaldistance,color=richter_scale))+
  geom_histogram(position = "identity", alpha = 0.6, binwidth = 0.5)+
  labs(title = "Depth vs Earthquake Magnitude",
       y = "horizontal diatnce") +
  theme_minimal()


p2=df_earthquake %>% filter(!is.na(horizontaldistance)) %>% 

    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  group_by(richter_scale) %>% 
  summarise(mean_hd=mean(horizontaldistance)) %>% 
  
  ggplot(aes(x=richter_scale,y=mean_hd))+
  geom_col(stat = 'identity')+
  labs(title = "Average horizontal distance per earthquake category",
       y = "horizontal diatnce") +
  theme_minimal()

p3=df_earthquake %>%
    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>%
  ggplot(aes(x=magnitude,y=horizontaldistance))+
  stat_cor(method = "pearson", label.x =6, label.y = 30)+
  geom_point()+
  geom_smooth()+
  labs(title = "Horizontal Distance vs Magnitude",x='magnitude',
       y = 'distance') +
  theme_minimal()




p4=df_earthquake %>%
    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>%
  ggplot(aes(x=magnitude,y=depth))+
  stat_cor(method = "pearson", label.x =8.5, label.y = 600)+
  geom_point()+
  geom_smooth()+
  labs(title = "Depth vs Magnitude",x='magnitude',
       y = 'depth') +
  theme_minimal()

(p1+p2)/(p3+p4)

```

```{r}
# Largest earthquakes in terms of magnitude
# Look at earthquakes in the 3rd quartile and higher
# third_qaurtile=summary(df_earthquake$magnitude)[[4]] # the mean
# 
# 
# gr_mag_3rd_quatile=df_earthquake %>% 
#   group_by(richter_scale) %>%
#   summarize(grouped_mean_mg=mean(magnitude))
# 
# left_join(df_earthquake,gr_mag_3rd_quatile, by ='richter_scale') %>% 
#   mutate( is_greater=
#   case_when(magnitude>=grouped_mean_mg~ 1,
#             magnitude<grouped_mean_mg~0)
#   
#   ) %>% 
#   select(id,richter_scale,magnitude,grouped_mean_mg,is_greater) %>% 
#   group_by(richter_scale) %>% 
#   summarize(total_ids=length(id),total_gr=sum(is_greater)  ) %>% 
#   mutate(percentage=(total_gr/total_ids))
  







```

```{r az_rms_mg,eval=TRUE, echo=FALSE}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
# Also explore relationship etween azimuthal gap,rms and earthqake magnitudes
library(reshape2)
corr_data=df_earthquake %>% select(azimuthalgap,rootmeansquare,magnitude,depth,deptherror,horizontaldistance,horizontalerror,magnitudeseismicstations) %>% 
  select(where(is.numeric)) %>% 
  cor(use = "pairwise.complete.obs")


corr_data_df=melt(corr_data)

corr_data_df %>% ggplot(aes(x=Var2, y=Var1, fill=value))+
geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  geom_text(aes(label = round(value, 2)), color = "black")+
  labs(title = "Feature Correlation Matrix",x='Feature',y='Feature')+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
  
```

```{r}
df_earthquake %>% select(id,time.x,time.y,dt_date_time,magnitude) %>% 
  mutate(year_=year(dt_date_time)) %>% 
  group_by(year_) %>% 
  summarize(nr_eq=n_distinct(id),avg_mg=mean(magnitude)) %>% 
  filter(nr_eq==max(nr_eq))

df_earthquake %>% mutate(year_=year(dt_date_time)) %>% 
select(id,time.x,time.y,dt_date_time,magnitude,year_) %>% 
  filter(is.na(year_))

```

# Temporal Data Analysis

```{r eq_per_dt, eval=TRUE, echo=FALSE}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 12

p1=df_earthquake %>% select(id,time.x,time.y,dt_date_time,magnitude) %>% 
  mutate(year_=year(dt_date_time)) %>% 
  group_by(year_) %>% 
  summarize(nr_eq=n_distinct(id),avg_mg=mean(magnitude)) %>% 
  arrange(year_) %>% 
   mutate(running_avg = rollmean(nr_eq, k =3, fill = NA, align = "right")) %>% 
  ggplot(aes(x=year_,y=nr_eq))+
  geom_col(stat = 'identity')+
  geom_line(aes(y =running_avg ), color = "blue")+
  geom_point(aes(y =running_avg ), color = "blue")+
  labs(title = "Number of Earthquakes per year",y='count',
       x= "year") +
  theme_minimal()
  
 p2=df_earthquake %>% select(id,time.x,time.y,date_final,magnitude) %>% 
  mutate(year_=year(date_final)) %>% 
  group_by(year_) %>% 
  summarize(nr_eq=n_distinct(id),avg_mg=mean(magnitude)) %>% 
  ggplot(aes(x=year_,y=avg_mg))+
  geom_line()+
  labs(title = "Yearly Average Earthquake Magnitude ",y='avg magnitude',
       x = "year") +
  theme_minimal()
 
 
p3=df_earthquake %>% select(id,time.x,time.y,date_final,depth) %>% 
  mutate(year_=year(date_final)) %>% 
  group_by(year_) %>% 
  summarize(nr_eq=n_distinct(id),avg_mg=mean(depth)) %>% 
  ggplot(aes(x=year_,y=avg_mg))+
  geom_line()+
  labs(title = "Yearly Average Earthquake Depth ",y='avg depth',
       x = "year") +
  theme_minimal()
 
 p4=df_earthquake %>% select(id,time.x,time.y,date_final,horizontaldistance) %>% 
  mutate(year_=year(date_final)) %>% 
  group_by(year_) %>% 
  summarize(nr_eq=n_distinct(id),avg_mg=mean(horizontaldistance)) %>% 
  ggplot(aes(x=year_,y=avg_mg))+
  geom_line()+
  labs(title = "Yearly Average Earthquake Horizontal Distance ",y='avg horizontal diatance',
       x = "year") +
  theme_minimal()
 
 
 (p2|p3)/p1

```

```{r}

# key questions:
# Look into time of day, do earthquakes tend to happen at certain times in the day,
#also is the a pattern in their magnitudes?
# Also look into how this breaks down per earthquake time
# Also explore relationship etween azimuthal gap,rms and earthqake magnitudes, create a distplot for this
# How has earthquake accuracy changed over time?
# Look into temporal correlations (spatial aswell, lookd into this at work)
# Extract country/location information from place variable in data- see see if we can group time series by location

# df_earthquake %>% select(place,latitude,longitude) %>% filter(!is.na(place))
# 
# names(query)
# 
# query %>% select(place,latitude,longitude) %>% filter(!is.na(place))
```

```{r,hr_day,eval=TRUE, echo=FALSE}

#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 13

library(ggridges)
# eq count and magnitude per time of day
p1=df_earthquake %>% select(dt_date_time,id,magnitude) %>% 
  mutate(hr_day=hour(dt_date_time)) %>% 
  group_by(hr_day) %>% 
  summarize(nr_eq=length(id),mean_mg=mean(magnitude)) %>% 
  ggplot(aes(x=hr_day,y=nr_eq))+
  geom_col(stat = 'identity')+
  labs(title = "Hourly Counts",y='count',
       x= "hour of day") +
  theme_minimal()

p2=df_earthquake %>% select(dt_date_time,id,magnitude) %>% 
  mutate(hr_day=hour(dt_date_time)) %>% 
  group_by(hr_day) %>% 
  summarize(nr_eq=length(id),mean_mg=mean(magnitude)) %>% 
  ggplot(aes(x=hr_day,y=mean_mg))+
  geom_line()+
  geom_point()+
  labs(title = "Hourly Mean Magnitudes",y='average magnitude',
       x= "hour of day") +
  theme_minimal()



p3=df_earthquake %>% select(dt_date_time,id,magnitude,richter_scale) %>%
  mutate(hr_day=hour(dt_date_time)) %>%
  mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  group_by(hr_day,richter_scale) %>%
  summarize(nr_eq=length(id),mean_mg=mean(magnitude)) %>%
  
  ggplot(aes(x=hr_day,y=nr_eq,fill=richter_scale))+
  geom_ridgeline(aes(height=nr_eq))+
  
  labs(title = "Counts Grouped by Category",y='count',
       x= "hour of day") +
  theme_minimal()


(p2|p1)/p3
```

Key find:

```{r,temp_corr}

# correlation between rms and magnitude - over time
# df_earthquake %>% 
#   mutate(year_=year(dt_date_time)) %>% 
#   group_by(year_) %>% 
#   summarise(avg_rms=mean(rms,rm.na=TRUE)) %>% 
#   ggplot(aes(x=year_,y=avg_rms))+
#   geom_line()+
#   labs(title = "Accuracy over time",y='root mean sqaured error',
#        x= "year") +
#   theme_minimal()

#df_earthquake %>% select(place,latitude,longitude) %>% filter(!is.na(place))


```

# Spatial Analysis

-   look into first filtering for

```{r}
df_earthquake %>% select(net) %>% filter(!is.na(net))

ids=query %>% filter(!is.na(net) & type_query=='earthquake') %>% distinct(id) %>% pull()

df_earthquake %>% filter(type=='Earthquake' )  %>% distinct(type_query)
```

```{r}
a=data.frame(num=1:10,val1=rep('num',times=10))
b=data.frame(num=2:20,val2=rep('vel',times=length(2:20)))


full_join(a,b)

```
