---
title: "EDA Assignmemnt2"
author: "Pamela Afful"
format: pdf
editor: visual

execute: 
  cache: true
  warning: false
  message: false
  echo: false
---

```{r, read_lib}
library(lubridate)
library(dplyr)
library(ggplot2)
#library(ggpubr)
```

# Data Wrangling

Read in the files

```{r,read_data,echo=FALSE,eval=TRUE}

query=read.csv("query.csv")
earth_quakes=read.csv("Earthquakes_1965_2016.csv")


```

Check Dataset Details

```{r,data_det,echo=FALSE,eval=TRUE}
# check query df details
dim(query)
dim(earth_quakes)
```

Convert time characters to date variables for both dataframes

Create truncated date variables, only including the date part

Query data

```{r}
names(query)
```

```{r,date_con,echo=FALSE,eval=FALSE}
# Change date columns to datetime
query$dt_date_time=ymd_hms(query$time)
query$date_final=date(query$dt_date_time)

```

Earthquakes data

```{r,date_con_2,echo=FALSE,eval=TRUE}
# Split data into 2 then put back together

earth_quakes_final=earth_quakes%>% mutate(dt_date_time= mdy_hms(paste(Date,Time)))
earth_quakes_final$date_final=date(earth_quakes_final$dt_date_time)

# earth_quakes_final$date_final=date(earth_quakes_final$dt_time)
head(earth_quakes_final %>% select(Date,Time,dt_date_time,date_final))
```

Transform column names to lower case for both dataframes

Also, remove the dot character, "." from the column names of the earthquake data and lower the case of the values in the id column for the earthquakes data

```{r,lower_col,echo=FALSE,eval=TRUE}
# lower the case of column names
colnames(query)=tolower(colnames(query))


# lower the case of the columns names in earth quake df and remove period character in column names
colnames(earth_quakes_final)=tolower(colnames(earth_quakes_final))
colnames(earth_quakes_final)=gsub("\\.","",colnames(earth_quakes_final))

# lower the case of the ID values
earth_quakes_final$id=tolower(earth_quakes_final$id)

# rename some columns in query dataframe
query=query %>% rename(magnitudetype=magtype) %>% 
  mutate(magnitudetype=toupper(magnitudetype),
         rootmeansquare=rms_query)



```

Finally join the dataframes on common column values

```{r,df_join,echo=FALSE,eval=TRUE}

# find common columns in both dataframes
eq_cols=names(earth_quakes_final)
query_cols=names(query)

join_cols=intersect(eq_cols,query_cols)
join_cols=join_cols[join_cols!='time']

df=left_join(earth_quakes_final,query, by=join_cols)

df%>% select(any_of(join_cols)) %>% head()
```

Now, filter for just earthquake datab

```{r,eq_fil,echo=FALSE,eval=TRUE}
# filter for earthqauke data
df_earthquake=df %>% filter(type=='Earthquake')


# Drop cols that have too many NA values or have repeated counterpart columns
names(df_earthquake)

keep_cols=c('date',"longitude","latitude","depthseismicstations", "magnitude",
            "magnitudeseismicstations","azimuthalgap","rootmeansquare",
            "id"
            )

df_earthquake %>% select(mag,magnitude)
int_ids=intersect(q_ids,eq_ids)

```

```{r}

names(earth_quakes_final)

earth_quakes_final %>% filter(id== 'us10007ntd') %>%
    select(id,magnitude,magnitudetype)

names(earth_quakes_final)
```

```{r}
names(query)

query %>% filter(id== 'us10007ntd') %>%
    select(id,mag,magnitudetype)
```

### Creating Magnitude Scale

```{r,mag_cat,echo=FALSE,eval=TRUE}
df_earthquake$richter_scale=df_earthquake %>% mutate(richter_scale=case_when(
                                                  #magnitudetype=='ML' &
                                                  magnitude>=1 & magnitude<=1.9~'Micro',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=2 & magnitude<=2.9~'Minor',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=3 & magnitude<=3.9~'Slight',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=4 & magnitude<=4.9~'Light',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=5 & magnitude<=5.99~'Moderate',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=6 & magnitude<=6.9~'Strong',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=7 & magnitude<=7.9~'Major',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=8 & magnitude<=8.9~'Great',
                                                  #magnitudetype=='ML' &
                                                  magnitude>=9 & magnitude<=9.9~'Extreme'
                                                  #magnitudetype!='ML'~'Other',
                                                  
                                                  
                                                  
                                                    
)
                         ) %>% 
  pull(richter_scale)


```

Find count/frequency of earth quake types per magnitude

```{r,bar_plot,echo=FALSE,eval=TRUE}

summary(df_earthquake$magnitude)


df_earthquake %>%  
  distinct(id,richter_scale) %>% 
  mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  ggplot(aes(x=richter_scale))+
  geom_bar()+
  labs(title = "Number of Earthquake Per Magnitude Category",
       x = "magnitude category") +
  theme_minimal()

df_earthquake %>%  
  select(magnitude,richter_scale) %>% 
  ggplot(aes(x=magnitude,color=richter_scale))+
  geom_histogram(position = "identity", alpha = 0.6, binwidth = 0.5)+
  labs(title = "Earth Quake Magnitudes Per Richter Scale Category (All Magnitude Types)",
       x = "magnitude") +
  theme_minimal()



  
```

```{r}
summary(df_earthquake$magnitude)

df_earthquake %>%  
  select(magnitude,richter_scale) %>% 
  ggplot(aes(y=magnitude))+
  geom_boxplot()+
  labs(title = "Distribution of Earthquake Magnitudes",
       y = "magnitude") +
  theme_minimal()


```

```{r}

df_earthquake %>%  
  select(magnitude,depth,richter_scale) %>% 
    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  ggplot(aes(x=richter_scale,y=depth))+
  geom_boxplot()+
  labs(title = "Depth vs Earthquake Category",
       y = 'earth quake depth') +
  theme_minimal()

```

```{r}
df_earthquake %>% filter(!is.na(horizontaldistance)) %>% 

    mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
  ggplot(aes(x=horizontaldistance,color=richter_scale))+
  geom_histogram(position = "identity", alpha = 0.6, binwidth = 0.5)+
  labs(title = "Depth vs Earthquake Magnitude",
       y = "horizontal diatnce") +
  theme_minimal()

# df_earthquake %>%  
#     mutate(richter_scale=factor(richter_scale,levels=c('Micro','Minor','Slight','Light','Moderate','Strong','Major','Great','Extreme'))) %>% 
#   ggplot(aes(x=horizontaldistance,y=magnitude))+
#   stat_cor(method = "pearson", label.x = 20, label.y = 8)+
#   geom_point()+
#   labs(title = "Depth vs Earthquake Category",
#        y = 'earth quake depth') +
#   theme_minimal()



```

```{r}
# Largest earthquakes in terms of magnitude
# Look at earthquakes in the 3rd quartile and higher
third_qaurtile=summary(df_earthquake$magnitude)[[5]]


df_earthquake %>% filter(magnitude>third_qaurtile) %>% 
  group_by(richter_scale) %>% 
  

```

```{r}
query %>% colnames()

earth_quakes_final %>% filter( !is.na(magnitudeseismicstations)) %>% 
  select (magnitudeseismicstations)
```

```{r}
q_ids=query %>% distinct(id) %>% pull()
eq_ids=earth_quakes_final %>% distinct(id) %>% pull()

int_ids=intersect(q_ids,eq_ids)

df_earthquake %>% filter(id %in%int_ids ) %>% 
  select(id,nst,magnitudeseismicstations,rootmeansquare,rms)

# earth_quakes_final %>% filter(id %in%int_ids ) %>% 
#   select(id,magnitudeseismicstations)



```

# Temporal Data Analysis

```{r}



```
