---
title: "EDA Assignmemnt2"
author: "Pamela Afful"
format: pdf
editor: visual
---

```{r, read_lib}
library(lubridate)
library(dplyr)
```

# Data Wrangling

Read in the files

```{r,read_data,echo=FALSE,eval=TRUE}

query=read.csv("query.csv")
earth_quakes=read.csv("Earthquakes_1965_2016.csv")


```

Check Dataset Details

```{r,data_det,echo=FALSE,eval=TRUE}
# check query df details
str(query)
str(earth_quakes)
```

Convert time characters to date variables for both dataframes

Query

```{r,date_con,echo=FALSE,eval=FALSE}
# Change date columns to datetime
query$dt_time=ymd_hms(query$time)


```

earthquakes

```{r,date_con_2,echo=FALSE,eval=TRUE}
# Split data into 2 then put back together

earth_quakes$dt_date=mdy(earth_quakes$Date)
earth_quakes_na=earth_quakes %>% filter( is.na(dt_date)) %>% 
  mutate(dt_date=ymd_hms(Date))

earth_quakes_non_na=earth_quakes %>% filter(!is.na(dt_date))

# concat back together
earth_quakes_final=rbind(earth_quakes_non_na,earth_quakes_non_na)

head(earth_quakes_final)
```

Create truncated date variables, only including the date part

```{r,truc_date,echo=FALSE,eval=TRUE}
# Create a truncated date function with no time elements 
earth_quakes_final$date_final=date(earth_quakes_final$dt_date)
earth_quakes_final %>% select(Date,dt_date,date_final)
```

```{r,trunc_date2,echo=FALSE,eval=TRUE}
# Do the same for the query csv
query$date_final=date(query$dt_time)
query %>% select(time, dt_time,date_final)
```

Transform column names to lower case for both dataframes

```{r,lower_col,echo=FALSE,eval=TRUE}
colnames(query)=tolower(colnames(query))
colnames(query)
```

Ensure to remove the period character in the earth quake data in addition to converting to lower case

```{r,lower_col2,echo=FALSE,eval=TRUE}
# lower the case of the columns names in earth quake df and remove period character in column names
colnames(earth_quakes_final)=tolower(colnames(earth_quakes_final))

colnames(earth_quakes_final)=gsub("\\.","",colnames(earth_quakes_final))

colnames(earth_quakes_final)
```

Lower the case of the values in the ID column of the earth quake dataset

```{r,lower_col3,echo=FALSE,eval=TRUE}
# lower the case of the id values in the id column  of the earthquakes dataframe
earth_quakes_final %>% select(id)
earth_quakes_final$id=tolower(earth_quakes_final$id)
earth_quakes_final %>% select(id)
```

Join the dataframes

```{r,df_join,echo=FALSE,eval=TRUE}
# find common columns in both dataframes
eq_cols=names(earth_quakes_final)
query_cols=names(query)

join_cols=intersect(eq_cols,query_cols)
join_cols

df_final=full_join(earth_quakes_final,query, by=join_cols)
head(df_final)
```

Filter for just earthquake data

```{r,eq_fil,echo=FALSE,eval=TRUE}
# filter for earthqauke data
df_earthquake=df_final %>% filter(type=='Earthquake')
df_earthquake
```
